Natural selection in the Health and Retirement Study

### 2023-09-18 hrs-selection.qmd
bugfixes

# TODO: use survey package
# here's some useful code https://github.com/ajdamico/asdfree/blob/master/27-hrs.Rmd
# TODO (related) move away from lists and towards a single datasource with an
# "ethnicity" column; results from broom::tidy can still be in lists?

@@

library(haven)
library(forcats)
library(dplyr, warn.conflicts = FALSE)
library(stringr)
library(purrr)
library(ggplot2)
library(tidyr)
library(santoku, warn.conflicts = FALSE) # masks tidyr::chop
library(survey)

@@

calc_income <- function (x, eth) {
  x <- x |> filter(ethnicity == eth)
  ages <- x |> select(matches("r\\d+agey_e"))
  income <- x |> select(matches("r\\d+iearn"))
  income <- rowMeans(income, na.rm = TRUE)
  income_r <- resid(lm(income ~ factor(x$rabyear), weights = x$weight))
  income_r <- income_r - min(income_r) + 1

@@

@@

rand <- rand |> left_join(rand_fam, 
                          by = join_by(hhid, pn),
                          unmatched = "drop",
                          relationship = "one-to-one")
rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + r10sbdied,
    agefb = r10agey_e - h10ageokid,
    agefb = ifelse(agefb < 12, NA_real_, agefb),
    agefb = chop_equally(agefb, 2, labels = c("Low", "High"))
  )

@@

# weighting
rand <- left_join(rand, trk_weights, by = join_by(hhid == HHID, pn == PN))
rand$weight <- ifelse(! is.na(rand$MBIOWGTR), rand$MBIOWGTR, rand$NBIOWGTR)
# compensate for fecund parents more likely to have child in survey:
rand$parent_weight <- rand$weight / (rand$r10livsib + 1)

@@

@@

The resulting sample contains `r nrow(rands$white)` genotyped white participants and `r nrow(rands$black)` genotyped black participants. Genotyping took place in 2006, 2008 and subsequent years. I use polygenic scores (PGS) precalculated by XXX. I discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. I also discard PGS for number of children ever born (but keep scores for age at first birth).

@@

```{r}
#| label: stat-regressions
#| cache: true

@@

  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
  mod_pgs[[eth]] <- map(form_pgs, 
                        \(x) svyglm(x, r_eth)
                        )

@@

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                              \(x) svyglm(x, r_eth)
                              )

@@

  form_pgs_sibs <- map(pgs, 
                       \(x) reformulate(c(x, pcs), response = "r10sibs"))
  mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
                             \(x) svyglm(x, r_parents_eth))

@@

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
        \(x) svyglm(x, r_eth)
        )

@@

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
        \(x) svyglm(x, r_eth)
        )

@@

  form_pgs_x_mar <- paste0(pgs, ":married") |>
    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
        \(x) svyglm(x, r_eth)
        )

@@

  form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
    map(\(x) reformulate(c(x, "agefb", pcs), response = "raevbrn"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
        \(x) svyglm(x, r_eth)
        )

@@

  cor_fert_edu_phen[[eth]] <- 
    svyglm(raevbrn ~ raedyrs, r_eth) |> 
    coef() |> 
    pluck("raedyrs")
}

@@

```
Appendix @fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group. Standard errors are large because of the relatively low sample sizes, and no scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`. But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores.

```{r}
#| label: stat-bootstraps
#| cache: true

@@

tails_boot_sibs <- map(cors_sibs_edu, 
                       \(x) x + c(-1.96, 1.96) * SE(x))
```

@@

The relationship is imprecisely estimated in the smaller black sample, but it is also steeper. Put another way, if we measured PGS in units of "years of associated education", estimated selection effects would be larger among blacks than whites. If this holds in the population, one reason may be that the correlation between phenotypic years of education and fertility is larger for blacks than for whites. In bivariate regressions, an extra year of education reduces fertility by `r cor_fert_edu_phen$white` for whites and `r cor_fert_edu_phen$black` for blacks [cf. @goldscheider1969minority; @johnson1979minority; @yang2003big]. In turn, the simplest explanation for this fact would be point 2 of the theory. Because black people on average have lower incomes and fewer years of education, and are more likely to be single parents, the fertility-wage tradeoff is especially acute for them.

@@

I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of siblings (including dead siblings) in 2010. I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so effect sizes are not comparable across the generations. Appendix @fig-siblings plots effects on number of siblings versus effects on years of education. The correlation is negative and significant for whites (correlation `r cors_sibs_edu$white`, bootstrapped 95% C.I. `r tails_boot_sibs$white[1]` to `r tails_boot_sibs$white[2]`). It is positive for blacks (correlation `r cors_sibs_edu$black`), although this is driven by two outlier PGS, which are scores for Alzheimer's with all SNPs included; without these, the correlation is negative.

@@

```{r}
#| label: stat-x-edu-bootstraps
#| cache: true

n_reps <- 199

cross_terms <- function (x) {
  cx <- coef(x)
  cx[str_detect(names(cx), ":")]
}

form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
form_pgs_x_inc <- paste0(pgs, ":income_med")  |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
form_pgs_x_mar <- paste0(pgs, ":married")  |>
    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))

coefs_boot_ed <- coefs_boot_inc <- coefs_boot_mar <- list()

for (eth in eths) {
  rand_short <- rand |> 
    filter(ethnicity == eth) |> 
    select(
      all_of(pgs), all_of(pcs), 
      raevbrn, raedyrs, weight, income_med, edu, married
    )

  coefs_boot_ed[[eth]] <- replicate(n_reps, {
    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
    map(form_pgs_x_ed, 
        \(x) lm(x, data = rand_r, weights = weight)
        ) |> 
      map(cross_terms) |> 
      unlist()
  })

  coefs_boot_inc[[eth]]  <- replicate(n_reps, {
    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
    map(form_pgs_x_inc, 
          \(x) lm(x, data = rand_r, weights = weight)
          ) |> 
      map(cross_terms) |> 
      unlist()
  })

  coefs_boot_mar[[eth]]  <- replicate(n_reps, {
    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
    map(form_pgs_x_mar, 
          \(x) lm(x, data = rand_r, weights = weight)
          ) |> 
      map(cross_terms) |> 
      unlist()
  })

  reorganize <- function (x) {
   x |> 
    t() |> 
    as.data.frame() |> 
    pivot_longer(everything(), names_to = c("pgs", "group"), 
                 names_sep = ":") |> 
    mutate(rep = 1:n_reps, .by = c(pgs, group)) |> 
    pivot_wider(id_cols = c(pgs, rep), 
                names_from = group, 
                values_from = value) 
  }

  coefs_boot_ed[[eth]] <- reorganize(coefs_boot_ed[[eth]])
  coefs_boot_inc[[eth]] <- reorganize(coefs_boot_inc[[eth]])
  coefs_boot_mar[[eth]] <- reorganize(coefs_boot_mar[[eth]])
}

boot_ci <- function (x) {
  names(x)[3:4] <- c("lo", "hi")
  x |> 
    mutate(
      abs_diff = abs(hi) - abs(lo)
    ) |> 
    summarize(
      mad = mean(abs_diff),
      .by = rep
    ) |> 
    pull(mad) |> 
    quantile(c(0.025, 0.975))
}

ci_inc_w <- boot_ci(coefs_boot_inc$white)
ci_inc_b <- boot_ci(coefs_boot_inc$black)
ci_ed_w <- boot_ci(coefs_boot_ed$white)
ci_ed_b <- boot_ci(coefs_boot_ed$black)
ci_mar_w <- boot_ci(coefs_boot_mar$white)
ci_mar_b <- boot_ci(coefs_boot_mar$black)

for (ci in list(ci_inc_w, ci_inc_b, ci_ed_w, ci_ed_b)) {
  if (prod(sign(ci)) != -1) {
    stop("effect size IS significantly different across some pair groups")
  }  
}
```

@fig-x-edu shows that for both whites and blacks, there is essentially no difference in the distribution of absolute effect sizes between respondents with more and less education. @fig-x-inc, on the other hand, shows that absolute effect sizes appear larger among lower-income respondents, for both whites and blacks. However, in bootstrap tests, the median difference in absolute effect sizes is not significantly different from 0, for either ethnic group. @fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses. Among blacks, effect sizes look similar across marital status, but among whites they are larger for unmarried respondents, and this is significant in bootstraps (95% C.I. for median difference `r ci_mar_w[1]` to `r ci_mar_w[2]`). @fig-x-agefb splits respondents by age at first birth (median within each ethnicity). There are no large differences between early and late parents. Note that the N is lower here due to missing values. TODO test

>>>>

@@

library(haven)
library(forcats)
library(dplyr, warn.conflicts = FALSE)
library(stringr)
library(purrr)
library(ggplot2)
library(tidyr)
library(santoku, warn.conflicts = FALSE) # masks tidyr::chop
 # masks dotchart:
suppressPackageStartupMessages(library(survey, warn.conflicts = FALSE))

@@

calc_income <- function (x, eth) {
  x <- x |> filter(ethnicity == eth)
  ages <- x |> select(matches("r\\d+agey_e"))
  income <- x |> select(matches("r\\d+iearn"))
  income <- rowMeans(income, na.rm = TRUE)
  mod <- lm(income ~ factor(x$rabyear), weights = x$weight, na.action =
              na.exclude)
  income_r <- resid(mod)
  income_r <- income_r - min(income_r, na.rm = TRUE) + 1

@@

# avoid warnings about zero weights
svyglm_quiet <- function (...) suppressWarnings(svyglm(...))

# for a confidence interval
significant <- function (x) {
  stopifnot(is.numeric(x), length(x) == 2)
  x[1] < 0 && x[2] > 0
}

insignificant <- Negate(significant)

@@

rand <- rand |> left_join(rand_fam, 
                          by = join_by(hhid, pn),
                          unmatched = "drop",
                          relationship = "one-to-one")

# weighting
rand <- left_join(rand, trk_weights, by = join_by(hhid == HHID, pn == PN))
rand$weight <- ifelse(is.na(rand$MBIOWGTR), rand$NBIOWGTR, rand$MBIOWGTR)
# compensate for fecund parents more likely to have child in survey.
# we use living sibs rather than including dead sibs too here. I think that's
# the correct approach:
rand$parent_weight <- rand$weight / (rand$r10livsib + 1)

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + r10sbdied + 1, # n. sibs including oneself!
    agefb = r10agey_e - h10ageokid,
    agefb = ifelse(agefb < 12, NA_real_, agefb),
    agefb = chop_equally(agefb, 2, labels = c("Low", "High"))
  )

@@

@@

n_white <- nrow(rand |> subset(ethnicity == "white"))
n_black <- nrow(rand |> subset(ethnicity == "black"))

# clean up cache so it doesn't get too big
rm(rand_orig, trk_orig, rands, rand_fam)
```

@@

The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants. Genotyping took place in 2006, 2008 and subsequent years. I use polygenic scores (PGS) precalculated by XXX. I discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. I also discard PGS for number of children ever born (but keep scores for age at first birth).

@@

```{r}
#| label: stat-regressions
#| cache: true
#| cache-lazy: false

@@

  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
  mod_pgs[[eth]] <- map(form_pgs, 
                        \(x) svyglm_quiet(x, r_eth)
                        )

@@

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                              \(x) svyglm_quiet(x, r_eth)
                              )

@@

  form_pgs_sibs <- map(pgs, 
                       \(x) reformulate(c(x, pcs), response = "r10sibs"))
  mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
                             \(x) svyglm_quiet(x, r_parents_eth))

@@

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
        \(x) svyglm_quiet(x, r_eth)
        )

@@

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
        \(x) svyglm_quiet(x, r_eth)
        )

@@

  form_pgs_x_mar <- paste0(pgs, ":married") |>
    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
        \(x) svyglm_quiet(x, r_eth)
        )

@@

  form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
    map(\(x) reformulate(c(x, "agefb", pcs), response = "raevbrn"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
        \(x) svyglm_quiet(x, r_eth)
        )

@@

  cor_fert_edu_phen[[eth]] <- 
    svyglm_quiet(raevbrn ~ raedyrs, r_eth) |> 
    coef() |> 
    pluck("raedyrs")
}

@@

Appendix @fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group. Standard errors are large because of the relatively low sample sizes, and no scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`. But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores.

```{r}
#| label: stat-bootstraps
#| cache: true
#| cache-lazy: false

@@

tails_boot_sibs <- map(cors_sibs_edu, 
                       \(x) x + c(-1.96, 1.96) * SE(x))

stopifnot(significant(tails_boot_fert$white))
stopifnot(insignificant(tails_boot_fert$black))
stopifnot(all(map_lgl(tails_boot_sibs, insignificant)))
```

@@

The relationship is imprecisely estimated in the smaller black sample, but it is also steeper. Put another way, if we measured PGS in units of "years of associated education", estimated selection effects would be larger among blacks than whites. If this holds in the population, one reason may be that the correlation between phenotypic years of education and fertility is larger for blacks than for whites. In bivariate regressions, an extra year of education reduces fertility by `r cor_fert_edu_phen$white` for whites and `r cor_fert_edu_phen$black` for blacks [cf. @goldscheider1969minority; @johnson1979minority; @yang2003big]. In turn, the simplest explanation for this fact would be point 2 of the theory. Because black people on average have lower incomes and fewer years of education, and are more likely to be single parents, the fertility-wage tradeoff is especially acute for them [cf. @maralani2013demography].

@@

I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of siblings (including dead siblings) in 2010. I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so effect sizes are not comparable across the generations. Appendix @fig-siblings plots effects on number of siblings versus effects on years of education. Correlations are insignificant for both ethnic groups.

@@

```{r}
#| label: stat-x-bootstraps
#| cache: true

n_reps_x <- 199

median_diff_effect <- function(w, data, term) {
  forms <- paste0(pgs, ":", term) |>
    map(\(x) reformulate(c(x, term, pcs), response = "raevbrn"))

  mods <- map(forms, \(x) lm(x, data, weights = w))
  cfs <- map(mods, coef)
  cross_terms <- map(cfs, \(x) x[str_detect(names(x), ":")])
  if(any(lengths(cross_terms) != 2)) stop("Extra cross terms")
  abs_cross_terms <- map(cross_terms, abs)
  diffs <- map_dbl(abs_cross_terms, diff) 
  median(diffs)
}

median_diff_ed <- median_diff_inc <- median_diff_mar <- median_diff_agefb <-
  list()

for (eth in eths) {
  rand_boot <- rand |> 
    subset(ethnicity == eth) |> 
    as.svrepdesign(type = "bootstrap", replicates = n_reps_x)

  median_diff_ed[[eth]] <- 
    withReplicates(rand_boot, theta = median_diff_effect, term = "edu")
  median_diff_inc[[eth]] <- 
    withReplicates(rand_boot, theta = median_diff_effect, term = "income_med")
  median_diff_mar[[eth]] <- 
    withReplicates(rand_boot, theta = median_diff_effect, term = "married")
  median_diff_agefb[[eth]] <-
    withReplicates(rand_boot, theta = median_diff_effect, term = "agefb")
}

confint_workaround <- function (x) {
  as.numeric(x) + c(-1.96, 1.96) * SE(x)
}

ci_ed <- map(median_diff_ed, confint_workaround)
ci_inc <- map(median_diff_inc, confint_workaround)
ci_mar <- map(median_diff_mar, confint_workaround)
ci_agefb <- map(median_diff_agefb, confint_workaround)

# TODO: is median_diff really a sensible test to run?

@fig-x-edu shows that for both whites and blacks, there is essentially no difference in the distribution of absolute effect sizes between respondents with more and less education. @fig-x-inc, on the other hand, shows that absolute effect sizes appear larger among lower-income respondents, for both whites and blacks. However, in bootstrap tests, the median difference in absolute effect sizes is not significantly different from 0, for either ethnic group. @fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses. Among blacks, effect sizes look similar across marital status, but among whites they are larger for unmarried respondents. @fig-x-agefb splits respondents by age at first birth (median within each ethnicity). There are no large differences between early and late parents. Note that the N is lower here due to missing values.
